---
title: "Trends in protected area types by period of establishment"
author: "Florent Bédécarrats, Stéphanie Carrière, Céline Fromont ¿et co-auteurs?"
format: 
  docx:
    message: false
    warning: false
    error: false
    echo: false
  html:
    message: false
    warning: false
    error: false
    code-fold: true
    toc: true
editor: visual
abstract: |
  We compute simple statistics on the database on Protected Areas Management 
  Evaluations (DB-PAME) to assess the prevalence of METT and RAPPAM methods among 
  other PAME methods. We show that METT and RAPPAM are very common in low and middle
  income countries, but rare in high income countries.
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

## Objective

In the framework of a paper in preparation [@fromont2023], we want to assess the prevalence of METT and RAPPAM methodologies among PAMEs. This document is designed to provide a verifiable and reproducible answer to this question. It is written in [quarto](https://quarto.org/) and uses R code. All the data processing can be viewed by clicking on the "code" buttons in the html version or in the code blocks in the Word or PDF versions.

```{r}
library(tidyverse, quietly = TRUE)
library(readxl)
library(gt)
dir.create("data")
```

## Data

We download the DB-PAME from the from [the protected planet portal](https://pame.protectedplanet.net/). It is the version updated in February 2022.

```{r}
pame_path <- "data/protectedplanet-pame-2023-FEB.csv"
if(!file.exists(pame_path)) {
  # Doesn not work, must be manual
  from = "https://www.protectedplanet.net/0e003033-2e0e-45a6-a312-2aaba2e5a3f9"
  download.file(from, destfile = pame_path, method = "curl")
}
pame <- read_csv(pame_path)
```

We download the World Bank country classification from the World bank data portal. The data comes in four classes: low income countries, low medium income countries and high medium income countries. The classification is available for all years since 1989, with some countries missing

```{r}

oghist_path <- "data/OGHIST.xls"

if (!file.exists(oghist_path)) {
  from = "https://databankfiles.worldbank.org/public/ddpext_download/site-content/OGHIST.xls"
  download.file(from, destfile = oghist_path, method = "curl")
}

oghist <- read_excel(oghist_path, sheet = "Country Analytical History", 
                     skip = 5) %>%
  rename(iso3 = 1, country_name = 2) %>%
  filter(!is.na(iso3))

missing <- oghist %>%
  pivot_longer(cols = matches("[0-9]{4}"), names_to = "year", values_to = "cat") %>%
  mutate(cat = ifelse(cat == "LM*", "LM", 
                      ifelse(cat == "..", NA, cat)))
test <- missing %>%
  group_by(year) %>%
  summarise(n = n(),
            na = sum(is.na(cat)),
            perc_na = round(na / n * 100, 2))
```

Status is missing for about a fourth of the countries at the beginning of the period (in particular countries from the ex soviet block). It is missing for less than 10 countries (\<4%) since 2010.

## Type of eval by country categories

We summarise the PAME production by income category of countries and by group of methods.

```{r}
#| label: tab-pame-method
#| tbl-cap: Prevalance of METT and RAPAM in PAME

# Include World Bank classification in PAME
pame <- pame %>%
  left_join(select(oghist, iso3, cat_wb = `2021`), by = "iso3") %>%
  mutate(method = ifelse(str_detect(methodology, "(METT)|(RAPPAM)"),
                         "METT or RAPPAM", "Other"))
  
method_top10 <- pame %>%
  filter(cat_wb %in% c("L", "LM")) %>%
  group_by(methodology) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  pluck("methodology")

# Summarise by method types

l_lm_um <- pame %>%
  filter(!is.na(cat_wb) & cat_wb != "H") %>% 
  group_by(cat_wb) %>%
  summarise(`Number of countries` = n_distinct(iso3),
            `Total number of assessments` = n(),
            `METT or RAPPAM (%)` = round(mean(method == "METT or RAPPAM") * 100, 2),
            `Other methods (%)` = round(mean(method == "Other") * 100, 2)) %>%
  rename(`County income category` = cat_wb)

lmic_hi <- pame %>%
  filter(!is.na(cat_wb)) %>% 
  mutate(cat_wb2 = ifelse(cat_wb == "H", "H", 
                          "LMIC")) %>%
  group_by(cat_wb2) %>%
  summarise(`Number of countries` = n_distinct(iso3),
            `Total number of assessments` = n(),
            `METT or RAPPAM (%)` = round(mean(method == "METT or RAPPAM") * 100, 2),
            `Other methods (%)` = round(mean(method == "Other") * 100, 2)) %>%
  rename(`County income category` = cat_wb2)
  
all <- bind_rows(l_lm_um, lmic_hi) %>%
  arrange(factor(`County income category`, levels = c("L", "LM", "UM", "LMIC", "H"))) %>% 
  mutate(`County income category` = case_when(
    `County income category` == "L" ~ "Low income",
    `County income category` == "LM" ~ "Lower middle income",
    `County income category` == "UM" ~ "Upper middle income",
    `County income category` == "LMIC" ~ "All low and middle income income",
    `County income category` == "H" ~ "Low income"))

gt(all)

```

From the results above, we can conclude that most of PAME registered in the DB-PAME occured in high income countries. We can also observe that METT and RAPPAM methods are dominant in LMIC, while they remain very marginal in developping countries.

## References
