---
title: "Trends in protected area types by period of establishment"
author: "Florent Bédécarrats, Stéphanie Carrière, Céline Fromont ¿et co-auteurs?"
format: 
  html:
    message: false
    warning: false
    error: false
    code-fold: true
    output-file: index.html
    toc: true
editor: visual
abstract: |
  We analyse the world database on protected areas to assess if the 
  relative importance of multipurpose protected areas has evolved compared to 
  strictly protected areas. We show very different trends whether we focus on the 
  number of entities or the spatial coverage of protected areas. During the last
  decades, there has been an increasing number of smaller strictly protected 
  areas, and a decreasing number of larger multipurpose protected areas. 

editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

# Objective

In the framework of a paper in preparation [@fromont2023], we want to know if the share of multi-purpose protected areas is increasing, compared to protected areas pursuing only conservation objectives. This document is designed to provide a verifiable and reproducible answer to this question. It is written in [quarto](https://quarto.org/) and uses R code. All the data processing can be viewed by clicking on the "code" buttons.

```{r}
# Install required libraries.
# Specified the used packages
required_packages <- c("tidyverse", "aws.s3", "wdpar", "sf", "gt", "rgee") 
# Identify which are missing
missing <- !(required_packages %in% installed.packages())
# Install those
if(any(missing)) install.packages(required_packages[missing], 
                                  repos = "https://cran.irsn.fr/")
# Load libraries
library(tidyverse)
library(aws.s3)
library(wdpar)
library(sf)
library(gt)
# Turn on/off scientific notations
options(scipen = 0) #999 for off
```

# Data

We use the extension wdpar to fetch the World Database on Protected Areas from [the protected planet portal](https://www.protectedplanet.net/en). It is the version updated in February 2022. We only keep the areas for which a spatial area is known, that is not including the protected area for which only one point in space is reported.

```{r}
#| output: false

# we also get the WDPA version from Google Earth Engine that includes areas 
save_object(object = "Replication_wolf/WDPA_no_geom_Feb2022.csv",
            file = "WDPA_no_geom_Feb2022.csv",
            bucket = "fbedecarrats",
            region = "")

# # For first download, use this code: 
# wdpa <- wdpa_fetch("global", download_dir = ".")
# # to save to S3 this one:
# put_object(file = "WDPA_Jan2023_Public.gdb.zip",
#            object = "Replication_wolf/WDPA_Jan2023_Public.gdb.zip",
#            bucket = "fbedecarrats",
#            region = "")
# # for more info on the use of S3 in this context, see: 
# # https://www.book.utilitr.org/01_r_insee/fiche_utiliser_rstudio_sspcloud

# Read the data
# wdpa <- wdpa_read("WDPA_Jan2023_Public.gdb.zip")
# This one was generated from Google Earth Engine to get the GIS area
wdpa_gee <- read_csv("WDPA_no_geom_Feb2022.csv")
```

As of this date, the WDPA includes information on `rnrow(wdpa_gee)` protected areas. A well-formatted description of the fields available can be found [on this webpage](https://developers.google.com/earth-engine/datasets/catalog/WCMC_WDPA_current_polygons).

# Methods

We keep only terrestrial protected areas and discard marine or coastal protected areas. Following Nelson et al. [-@nelson2011], we classify as strict protection the protected areas with a status corresponding to IUCN classes I through IV, and as multipurpose the protected areas with status corresponding to IUCN classes V and VI. 

For now, we have not figured out how to specify indigenous stewardship, so we leave it among the Unknown.

```{r}
# Drop marine, compute decade
wdpa2 <- wdpa_gee %>%
  filter(MARINE == "0") %>% #0: 100% terrestrial, 1: both, 2: 100% marine
  mutate(decade = STATUS_YR - (STATUS_YR %% 10),
         decade = case_when(
           decade == 0 ~ "Unknown",
           decade >0 & decade <= 1950 ~ "Before 1960",
           .default = paste(decade, decade+9, sep = "-")),
         type = ifelse(IUCN_CAT %in% c("Ia", "Ib", "II", "III", "IV"), "Strict",
                       ifelse(IUCN_CAT %in% c("V", "VI"), "Nonstrict", 
                              "Unknown")))
# Organize the order of display of decades, starting with Unnkown, then Before
decades <- wdpa2 %>%
  pluck("decade") %>%
  unique() %>%
  sort()
order_decades <- c("Unknown", "Before 1960", str_subset(decades, "^[0-9]")) 
wdpa2 <- wdpa2 %>%
  mutate(decade = factor(decade, levels = order_decades))
```

After this filtering, we have `r nrow(wdpa2)` protected areas for the analysis.

# Results

## Relative importance of status in number of PAs

@fig-n-pas represents the evolution of strict vs. multipurpose categories according to the year of assignment of the conservation status of PAs.

```{r fig_n_pas}
#| label: fig-n-pas
#| fig-cap: "Number of terrestrial protected areas by status and decade of creation"

# Descriptive statistics
patype_decade <- wdpa2 %>%
  group_by(decade, type) %>%
  summarise(n = n(),
            area = sum(GIS_AREA, na.rm = TRUE),
            mean_area = mean(GIS_AREA, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(type = factor(type, levels = c("Strict", "Nonstrict", "Unknown")))
patype_decade %>%
  ggplot(aes(x = decade, y = n, fill = type)) +
  geom_col() +
  xlab("Decade of PA status") + 
  ylab("Number of PAs receiving status") +
  labs(fill = "Category of status") + 
  scale_fill_discrete(
    labels = c("Strict" = "Strict (IUCN I to IV)",
               "Nonstrict" = "Multipurpose (IUCN V and VI)")) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

The @fig-n-pas shows that the total number of PAs has sharply increased since 2000. However, the relative importance of multipurpose conservation status doesn't seem to increase. 

```{r table_n_pas}
#| label: tab-n-pas
#| tbl-cap: "Area of PAs by status category and decade of creation"

# Stats in number of PAs
prop_patype_decade <- patype_decade %>%
  select(-area, -mean_area) %>%
  pivot_wider(names_from = type, values_from = n) %>%
  mutate(total = Nonstrict + Strict + Unknown,
         known_status = Nonstrict + Strict,
         perc_nonstrict = round(Nonstrict / known_status * 100, 1))
# Let's re-order colunms
prop_patype_decade %>%
  rename(`Decade of status assignment` = decade,
         `Total number of PAs created` = total,
         `Number of PAs whith a known status` = known_status,
         `% of multipurpose among known status` = perc_nonstrict) %>%
  gt()
  
  
```

According to the world database on protected areas, the proportion of new protected areas with a multipurpose status has reduced, from `r filter(prop_patype_decade, decade == "1960-1969") %>% pluck("perc_nonstrict")`% in the 1960s to `r filter(prop_patype_decade, decade == "2010-2019") %>% pluck("perc_nonstrict")`% in the 2010s. It has been even lower (`r filter(prop_patype_decade, decade == "2020-2029") %>% pluck("perc_nonstrict")`%) in the first years of the current decade. 

We must, however, strikeout that the decade of creation is unknown for `r filter(prop_patype_decade, decade == "Unknown") %>% pluck("total")` out of `r nrow(wdpa2)` protected area ( `r round(filter(prop_patype_decade, decade == "Unknown") %>% pluck("total")/ nrow(wdpa2)*100, 1)`%). There is a large proportion of multipurpose status among the protected areas for which the creation date is unknown (`r filter(prop_patype_decade, decade == "Unknown") %>% pluck("perc_nonstrict")`%). The status of the protected area is unknown for `r sum(prop_patype_decade$Unknown)` protected areas out of `r nrow(wdpa2)` protected areas (`r round(sum(prop_patype_decade$Unknown)/nrow(wdpa2)*100, 1)`%). 

## Relative importance of status in PA spatial extent

We now perform the same analysis, but instead of counting the number of PAs, we sum their area. The area newly covered by type of status is represented in @fig-area-pas.

```{r fig_area_pas}
#| label: fig-area-pas
#| fig-cap: "Area of terrestrial protected areas by status and decade of creation"

patype_decade %>%
  ggplot(aes(x = decade, y = area, fill = type)) +
  geom_col() +
  xlab("Decade of PA status") + 
  ylab(bquote('Area of PAs receiving status in km'^2)) +
  labs(fill = "Category of status") + 
  scale_fill_discrete(
    labels = c("Strict" = "Strict (IUCN I to IV)",
               "Nonstrict" = "Multipurpose (IUCN V and VI)")) + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

The trend that appears in @fig-area-pas is quite different than the one we saw in @fig-n-pas: when accounting for their total areas, multi-purpose protected areas seem to represent an increasing proportion among newly created protected areas.

We now compute the precise area estimates and their relative importance.

```{r table_area_pas}
#| label: tab-area-pas
#| tbl-cap: "Area of terrestrial protected areas by status and decade of creation"

# Stats in number of PAs
prop_patype_decade_area <- patype_decade %>%
  select(-n, -mean_area) %>%
  pivot_wider(names_from = type, values_from = area) %>%
  mutate(total = Nonstrict + Strict + Unknown,
         known_status = Nonstrict + Strict,
         perc_nonstrict = round(Nonstrict / known_status * 100, 1))
# Let's re-order colunms
prop_patype_decade_area %>%
  rename(`Decade of status assignment` = decade,
         `Total area of PAs created` = total,
         `Area of PAs whith a known status` = known_status,
         `% of multipurpose among known status` = perc_nonstrict) %>%
  gt()
```

According to the world database on protected areas, the total spatial extent of new protected areas with a multipurpose status has increased, from `r filter(prop_patype_decade_area, decade == "1960-1969") %>% pluck("perc_nonstrict")`% in the 1960s to `r filter(prop_patype_decade_area, decade == "2000-2009") %>% pluck("perc_nonstrict")`% in the 2000s and slightly decreased in the 2010s to `r filter(prop_patype_decade_area, decade == "2010-2019") %>% pluck("perc_nonstrict")`%. 

## Average size of protected areas

The two previous figures suggest very different trends in terms of average size of protected areas depending on their status. We verify this with @fig-avg-area.

```{r}
#| label: fig-avg-area
#| fig-cap: "Average area of terrestrial protected areas by status and decade of creation"

patype_decade %>%
  ggplot(aes(x = decade, y = mean_area, group = type, col = type)) +
  geom_point() +
  geom_line() +
  xlab("Decade of PA status") + 
  ylab(bquote('Average area of PAs in km'^2)) +
  labs(col = "Category of status") + 
  scale_fill_discrete(
    labels = c("Strict" = "Strict (IUCN I to IV)",
               "Nonstrict" = "Multipurpose (IUCN V and VI)")) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

@fig-avg-area shows that the area of PAs with a strict status has decreased over time, while the area of PAs with a multipurpose status has increased over time.

# Conclusion

The information available in the world database on protected areas indicates that **in terms of number of administrative entities**, the proportion protected areas with a multi-purpose status tends to decrease over the last decades. However, **in terms of spatial extent of these entities**,the proportion of protected area with a multi-purpose status tends to increase over the same period. In other words, countries create an increasing number of smaller strictly protected areas, and they create a decreasing number of larger multi-purpose protected areas. 

This conclusion, however, could be undermined by the substantial proportion of missing information on the creation date and status of protected areas in the database of reference. 

# References
