---
title: "Trends in protected area types by period of establishment"
author: "Florent Bédécarrats"
format: html
editor: visual
abstract: |
  This is a quick analysis aiming at computing the evolution of protected area 
  IUCN types depending on the period of creation of the protected areas. 

---


# Objective

We argue that the share of multi-purpose protected areas is increasing, compared to protected areas pursuing only conservation objectives.
This document is designed to be verificable and reproducible. It is written in quarto and uses R code. All the data processing can be viewed by clicking on the "code" buttons.

```{r}
# Install required libraries.

```


# Data

We use the extension wdpar to fetch the World Database on Protected Areas from [the protected planet portal](https://www.protectedplanet.net/en). It is the version updated in January 2022.

```{r}


df_PAs <- 
  aws.s3::s3read_using(
    FUN = data.table::fread,
    # Mettre les options de FUN ici
    object = "/FILOSOFI/2016/FILOSOFI_COM.csv",
    bucket = "donnees-insee",
    opts = list("region" = "")
  )
```


# Methods



# Results

# Discussion

